P3692R0: How to Avoid OOTA Without Really Trying

### 2025-06-11 paper/ootavacuous.tex
Reformulate in terms of preserving dependencies
Reformulate the paper's conception of semantic dependency relative to
a particular compiler.  In this new version, the text talks instead
about compiled code preserving syntactic dependencies.

In the original paper, perhaps the most questionable aspect was the
unspoken issue of whether the definition of semantic dependency did an
adequate job of capturing the informal notion.  In the new version,
this issue translates into an explicit Dependency Preservation thesis,
stating that semantic dependencies are always preserved.

In addition to the vir

	We focus on shared-memory programs and C++ implementations based
	on traditional compilers and hardware, including CPUs and GPGPUs,
	thus enabling analysis of semantic dependencies and OOTA behaviors
	by means of existing hardware models.
	We show that these models' constraints prevent OOTA cycles from
	occurring in undefined-behavior-free C++ programs running on
	such implementations, provided the cycles involve only volatile
	atomics.
	We accommodate nonvolatile atomics by defining ``quasi-volatile''
	behavior and show that existing implementations doing per-thread
	analysis and optimization not only avoid OOTA, but also avoid
	incorrectly evaluating arithmetic expressions involving atomic
	accesses.

	It follows that enforcing OOTA avoidance requires no special
	action from implementations and no changes to user code or
	to the standard.

@@

Unfortunately, the standard does not explain what the phrase ``depend on''
in the quotation above really means.
It is generally taken to mean \emph{semantic dependency},
a somewhat vague concept.
Roughly speaking, there is a semantic dependency from a given load to
a given store in the same thread when
\emph{all other things being equal, a change in the
value loaded can change the value stored, change the target address of
the store, or prevent the store from occurring at all.}
In Listing~\ref{lst:Simple OOTA}, for example,
there is a semantic dependency from line~1 to line~2 in both T0 and T1.
The semantic dependencies here are trivial,
because the values stored simply \emph{are} the values that were loaded.
(A \emph{syntactic} dependency, by contrast, exists whenever the value
of the load appears in the source code as part of the computation
leading up to the store, whether it affects the store or not.
\emph{@@@ Reference?}
Only semantic dependencies contribute to OOTA.)

Since real-world CPUs cannot store something
until they have determined its value,\footnote{
	Another way of saying this is that real-world CPUs do not
	make their stores visible to other CPUs until those stores
	are no longer speculative.
	A similar restriction applies to compiler-based speculation.}
the stores on line~2 cannot take place until T0's and T1's CPUs
know what values are returned by the loads on line~1.
Thus the hardware orders these stores after their corresponding loads,
and this ordering prevents the OOTA result.

@@

The inability of rfi links to act as temporal constraints is not the
only, or even the main, weakness in the naive argument against
OOTA cycles.
The primary difficulty lies in the fact that the code transformations
performed by optimizing compilers can destroy dependencies, including
semantic ones (depending on one's definition).
That is, even when the potential for a dependency exists in the
source code for a thread, there might be no dependency in the machine
code produced by a compiler.
There would then be no constraint forcing the implementation to execute
the thread's store later than the load it supposedly depends on,
and thus no impediment to the occurrence of an OOTA cycle.
We will see examples of this destruction later on.

@@

\section{A Definition of Semantic Dependency}
\label{sec:A Definition of Semantic Dependency}

Semantic dependency is a notoriously difficult concept to define
rigorously and precisely.
A large part of the reason is because it was never a completely clear
concept to begin with, especially when there are multiple accesses to
the variables involved.
In this section we will stick our necks out by offering just such a
definition.
No doubt many people will object to it for various reasons, but we
nevertheless hope it will help move the discussion forward.

The definition given below is applicable only to C++ implementations
that treat all atomic objects as though they are volatile or
quasi volatile.
(For compilers that perform only per-thread analysis, not global
analysis, quasi volatile is sufficient.)
In this setting we can relate abstract and hardware executions by
means of the map of accesses described in
Section~\ref{sec:Volatile and Quasi Volatile Accesses}.
The key insight is that this allows us to consider semantic
dependencies at the level of the machine code, where they are much
more tractable.
Using this approach, we demonstrate that OOTA cycles cannot form in the
C++ implementations addressed by this paper.

@@

We begin by recognizing that semantic dependencies are relative to a
particular execution and a particular implementation,
as described in Section~\ref{sec:Properties of Semantic Dependencies}.
The same source code may or may not contain a semantic dependency,
according to the details of the execution in question and the
machine code produced by the compiler.
For this reason we will characterize semantic dependencies in a given
abstract execution realized by a given hardware execution.
(While it is possible to argue about semantic dependencies in abstract
executions that have no hardware realizations, doing so seems pointless.)

Let $A$ be an abstract execution of some program $P$ containing a
thread $T$, and let $H$ be a hardware execution realizing $A$.
Let $W$ in $T$ be a store to an atomic object, and let $\{R_0,$
\ldots,~$R_n\}$ in $T$ be a set of loads from atomic objects on which
$W$ might or might not depend.
We can dispose of one case immediately: If $W$ is omitted in $H$ then
the issue of semantic dependency is moot.
You can give either answer since it will have no effect.
Therefore we'll assume that $W$ is not omitted.
Then:
\begin{quote}
There is a semantic dependency from $\{R_i\}$ to $W$ in $A$ and $H$,
relative to the compiler used to produce $H$, if there is another
abstract execution $B$ realized by hardware execution $G$ under the
same compiler that together \emph{witness the semantic dependency}.
\end{quote}

@@

In this context our definition of semantic dependency is essentially
the same as before.
Since we can no longer expect the machine code for a thread to be the
same regardless of the program it belongs to, the program $Q$ in the
earlier definition (of which $B$ and $G$ are executions) must be
$P$ or its volatilized form.
However, we do now allow the possibility that the executions $B$ and
$G$ take place in a nonbenign environment.
Aside from these minor adjustments, the definition remains unchanged.

@@

Of course we want to check that our definition of semantic dependency
satisfies the Fundamental Property of
Section~\ref{sec:The Fundamental Property of Semantic Dependencies}.
Given the information we have already presented, the demonstration is
easy.

Suppose we have $W$, $\{R_i\}$, $A$, and $H$ as in the definition.
The Fundamental Property assumes that $W$ is not omitted in $H$, so
there is an abstract execution $B$ with hardware realization $G$
witnessing the semantic dependency of the store $W$ on the loads
$\{R_i\}$ in $A$ and $H$.
We must show that some $R_i$ is not omitted and $R'_i$ is ordered
before $W'$ in $H$.
The proof splits into three cases.

First case: $W'$ lies in the initial period of $T$ in $H$.
During the initial period of the hardware executions, $H$ and $G$
behave identically and therefore $W'$ performs the same write in
both.
This contradicts the fact that $B$ and $G$ witness the semantic
dependency.

Second case: $W'$ lies in the common period of $T$ in $H$.
Since the action of $W'$ in $H$ is different from its action in $G$,
at least one of its inputs must differ between the two hardware
executions.
Therefore the source instruction for that input must behave
differently, and so must one of its sources, going back until we reach
a load instruction that obtains differing values in $H$ and $G$.
Then $W'$ depends on this load and so is ordered after it.
And since the load must lie in the common period of $H$, by the
definition of semantic dependency it must either be one of the
$R'_i$ or be ordered after one of them.
Therefore $W'$ is ordered after one of the $R'_i$ in $H$,
which certainly means that $R_i$ is not omitted.

@@

A corollary of this result is that if an implementation's compiler
either
(1)~uses per-thread analysis and treats atomic objects as quasi
volatile, or
(2)~uses global analysis and treats atomic objects as volatile,
then programs produced by that implementation will never exhibit OOTA.
Thus the implementation will automatically comply with full C++, even
if it was designed only to comply with loose C++.

@@

Here we consider some general questions related to our definition of
semantic dependency.
Additional information is available in P3064R2 (``How to Avoid OOTA Without
Really Trying'')~\cite{PaulEMcKenney2023P3046R2/OOTAwithoutTrying}.

\subsection{Relative vs.\ Absolute Dependency}
\label{sec:Relative vs. Absolute Dependency}

One drawback of our definition of semantic dependency is that it is only
relative to a specific compiler or implementation.
This may strike some readers (especially those implementing tooling)
as wishy-washy and avoiding the real problem, in that a given piece
of code should either contain or not contain a semantic dependency,
independent of the implementation used to run it.

We can address this drawback by defining an \emph{absolute semantic
dependency} as one that is present relative to any valid loose C++
implementation, past, present, or future, real or imagined.
Of course this notion has its own problems, including that it is
extremely nonconstructive and impossible to apply in practice.
However it may be the best we can do with our current understanding
of computing systems.

There is one thing we can definitely state:
Programs produced by an implementation of the right sort will never
exhibit absolute OOTA (that is, an OOTA cycle in which all the
semantic dependencies are absolute).
This is simply because an absolute semantic dependency is {\it
a fortiori\/} a semantic dependency relative to the compiler in use.

But in fact we have shown more than this:
Programs will never exhibit an OOTA cycle relative to the compiler
used to build them, even when that cycle is not absolute.
In this sense we have gone beyond the requirement of full C++.

@@

In principle we don't need to require global-analysis compilers to
treat atomic objects as volatile; our results would still hold if they
merely treated them as quasi volatile.
We chose not to do this because it would violate our intuitions about
semantic dependencies.

For example, consider the Simple OOTA program shown in
Listing~\ref{lst:Simple OOTA}.
A loose C++ compiler using global analysis and treating \co{x} and \co{y} as
quasi-volatile objects could omit the two loads, replacing them in the
machine code with simple assignments ``\co{r1 = 42}'' and
``\co{r2 = 42}'', as shown in
Section~\ref{sec:Global Optimization Can Destroy Dependencies}.
This would be a valid transformation, and the resulting behavior would
not count as OOTA according to our definition because the
dependencies in T0 and T1 would not be semantic.

To see why not, recall that a semantic dependency must have a witness,
another execution in which the store acts differently.
But this transformed program has no other hardware executions; every time it
runs it will store 42 to both \co{x} and \co{y}.
(Keep in mind also that since the atomic objects are not treated as
volatile, they are not subject to unspecified interference when the
program runs in a nonbenign environment.)

This unintuitive behavior could not occur if the two loads were not
omitted.
In fact, the definition of semantic dependency might remain
perfectly acceptable if the requirement for global-analysis compilers
were weakened, if the compiler treated atomic objects as quasi
volatile and in addition was not allowed to omit accesses to them.
This is a possible topic for future research.

@@

Part of our demonstration of the Fundamental Property of semantic
dependencies relies on the fact, stated in
Section~\ref{sec:Abstract Executions},
that an abstract execution of a thread is entirely determined by the
values obtained for its loads.
But when we compare abstract executions of the same thread in two
different programs, this may no longer be entirely true owing to the
effect of differing memory layouts.

@@

To rule out such pathological counterexamples we should require that
in a witness to a semantic dependency, the addresses of all the
objects and functions referred to in the thread $T$ are the same as
in the original execution.
This is a very technical restriction but there are occasions when
the issue might realistically arise, such as when computing a hash
value based on an object's address.

@@

To demonstrate the semantic dependency in the original code,
a suitable witness would have
to include a hardware realization of the abstract execution in which
\co{r1} and \co{r2} are zero and \co{r3} and \co{r4} are one.
But there are no hardware realizations of this execution with the
machine code indicated above!
Since \co{r1} is different from both \co{r3} and \co{r4}, the \co{r2}
load will be merged with the \co{r3} load and so \co{r2} will
necessarily be one, not zero.
Thus, relative to this compiler the example does not contain a
semantic dependency.

Although this is unexpected, and while we don't recommend this sort of
code transformation, we cannot say that this conclusion is definitely
wrong, because one's intuitive notions of semantic dependency are not
always clear in cases involving multiple loads from the same variable.
% Can we instead say that transforming the code might require that the
% witness also be transformed?
In the unlikely event that this transformation proves useful, there
will be a community process that determines whether the value of this
transformation is worth the added conceptual load on developers.
Until that time, pure mathematics cannot help us simply because it is
impossible to say whether or not there is a semantic dependency from
\co{y} to \co{z} without reference to a specific implementation.

@@

Additional future work could:
(1)~Extend these results to some classes of interpreters and just-in-time
compilers (JITs) on
one hand, and to link-time optimization (LTO) on the other;
(2)~Relax quasi-volatile semantics to permit limited load speculation
(for example, hoisting loads out of the bodies of conditionals);
(3)~Consider various classes of global analysis, such as
demonstrating that a given expression will always evaluate to a constant;
(4)~Examine the possibility of ``flattening'' optimizations that combine
multiple threads into one;
(5)~Explore cases in which semantic dependencies must be absolute,
not relative to any particular compiler;
(6)~Include the effects of input and output or other operations which
can vary from one execution to another; and
(7)~Delineate more precisely the limits of permissible behavior for
quasi-volatile object accesses.
This last item may well shed additional light on the dangers of
optimizations involving non-volatile atomic operations beyond the cautions
outlined in
Section~\ref{sec:Inventing Atomic Loads Can Cause Errors and
Destroy Semantic Dependencies}.

@@

This paper focused solely on compilers, primarily those that do only
per-thread analysis and optimization.
It introduced the notion of quasi-volatile object accesses and gave
a definition for semantic dependency that is relative to a particular
C++ implementation.
It formulated a Fundamental Property of semantic dependency and proved
that implementations subject to some very mild restrictions on how
they treat atomic object accesses do have this property.
Finally, it demonstrated that satisfying the Fundamental Property
guarantees an implementation cannot give rise to OOTA executions.

>>>>

	We focus on shared-memory programs and C++ implementations based
	on traditional compilers and hardware, including CPUs and GPGPUs,
	thus enabling analysis of semantic dependencies and OOTA behaviors
	by means of existing hardware models.
	We show that these models' constraints prevent OOTA cycles from
	occurring in undefined-behavior-free C++ programs running on
	such implementations, provided the cycles involve only volatile
	atomics.
	We accommodate nonvolatile atomics by defining ``quasi-volatile''
	behavior and show that for existing implementations doing per-thread
	analysis and optimization, this behavior not only avoids OOTA
	but also is necessary to avoid
	incorrectly evaluating arithmetic expressions involving atomic
	accesses.

	It follows that enforcing OOTA avoidance requires no special
	action from implementations and no changes to user code.

@@

Unfortunately, the standard does not explain what the phrase ``depend on''
in the quotation above really means.
It is generally taken to mean \emph{semantic dependency},
a somewhat vague concept.
Roughly speaking, there is a semantic dependency from a given load to
a given store in the same thread when
\emph{all other things being equal, a change in the
value loaded can change the value stored, change the target address of
the store, or prevent the store from occurring at all.}
In Listing~\ref{lst:Simple OOTA}, for example,
there is a semantic dependency from line~1 to line~2 in both T0 and T1.
The semantic dependencies here are trivial,
because the values stored simply \emph{are} the values that were loaded.

A \emph{syntactic} dependency, by contrast, exists whenever the value
of the load appears in the source code as part of the computation
leading up to the store, whether it affects the store or not.
\emph{@@@ Reference?}
Although we will not give a precise definition of syntactic dependency
here, doing so presents no fundamental difficulties.
In fact, the C++ standard covers much of the same ground in its definition
of ``carries a dependency'' in 6.9.2.2p7 (\co{[Data races]}).
Of course, all semantic dependencies are also syntactic.
Note, however, that only semantic dependencies contribute to OOTA.

Since real-world CPUs cannot store something
until they have determined its value,\footnote{
	Another way of saying this is that real-world CPUs do not
	make their stores visible to other CPUs until those stores
	are no longer speculative.
	A similar restriction applies to compiler-based speculation.}
the stores on line~2 of Listing~\ref{lst:Simple OOTA}
cannot take place until T0's and T1's CPUs
know what values are returned by the loads on line~1.
Thus the hardware orders these stores after their corresponding loads,
and this ordering prevents the OOTA result.

@@

The inability of rfi links to act as temporal constraints is not the
only, or even the main, weakness in the naive argument against
OOTA cycles.
The primary difficulty lies in the fact that the code transformations
performed by optimizing compilers can destroy syntactic dependencies,
possibly even semantic ones (depending on one's definition).
That is, even when a syntactic dependency exists in the
source code for a thread, there might be no dependency in the machine
code produced by a compiler.
There would then be no constraint forcing the implementation to execute
the thread's store later than the load it supposedly depends on,
and thus no impediment to the occurrence of an OOTA cycle.
We will see examples of this destruction later on.

@@

\section{Preservation of Dependencies}
\label{sec:Preservation of Dependencies}

Semantic dependency is a notoriously difficult concept to define
rigorously and precisely.
A large part of the reason is because it was never a completely clear
concept to begin with, especially when there are multiple accesses to
the variables involved.
In this section we will bypass the difficulty by instead defining
when an execution of compiled code \emph{preserves a syntactic dependency}.

In addition, we propose the following Dependency Preservation (DP) thesis:
\begin{quote}
	If a syntactic dependency is semantic then it is preserved
	by all compilers for which the definition below applies.
\end{quote}
As with the Church-Turing thesis in computability theory,
our DP thesis is not susceptible of formal proof because
the main concept it deals with (semantic dependency)
does not have a formal definition.
Nevertheless, we hope that readers will agree our definition of
dependency preservation captures the informal notion of
semantic dependency sufficiently well to justify the thesis.
It will be all we need to rule out the possibility of OOTA.

The definition given below is applicable only to C++ implementations
that treat all atomic objects as though they are volatile or
quasi volatile.
(For compilers that perform only per-thread analysis, not global
analysis, quasi volatile is sufficient.)
In this setting we can relate abstract and hardware executions by
means of the map of accesses described in
Section~\ref{sec:Volatile and Quasi Volatile Accesses}.
The key insight is that this allows us to consider
dependencies at the level of the machine code, where they are much
more tractable.

@@

We begin by recognizing that preservation of dependencies is relative to a
particular execution,
as described in Section~\ref{sec:Properties of Semantic Dependencies}.
A syntactic dependency present in the source code may or may not be
preserved in the machine code produced by a compiler,
according to the details of the execution in question.
For this reason we will characterize preservation of dependencies in a given
abstract execution realized by a given hardware execution.
(While it is possible to argue about preservation of dependencies in abstract
executions that have no hardware realizations, doing so seems pointless.)

Let $A$ be an abstract execution of some program $P$ containing a
thread $T$, and let $H$ be a hardware execution realizing $A$.
Let $W$ in $T$ be a store to an atomic object, and let $\{R_0,$
\ldots,~$R_n\}$ in $T$ be a set of loads from atomic objects on which
$W$ has a syntactic dependency.
We can dispose of one case immediately: If $W$ is omitted in $H$ then
the issue of dependency preservation is moot.
You can give either answer since it will have no effect.
Therefore we'll assume that $W$ is not omitted.
Then:
\begin{quote}
The dependency from $\{R_i\}$ to $W$ in $A$ is preserved in $H$
if there is another
abstract execution $B$ realized by hardware execution $G$
(of machine code produced by the same compiler as $H$)
that together \emph{witness the dependency}.
\end{quote}

@@

In this context our definition of dependency preservation is essentially
the same as before.
Since we can no longer expect the machine code for a thread to be the
same regardless of the program it belongs to, the program $Q$ in the
earlier definition (of which $B$ and $G$ are executions) must be
$P$ or its volatilized form.
However, we do now allow the possibility that the executions $B$ and
$G$ take place in a nonbenign environment.
Aside from these minor adjustments, the definition remains unchanged.

@@

Of course we want to check that whenever a dependency is preserved,
it satisfies the Fundamental Property of
Section~\ref{sec:The Fundamental Property of Semantic Dependencies}.
Given the information we have already presented, the demonstration is
easy.

Suppose we have $W$, $\{R_i\}$, $A$, and $H$ as in the definition.
The Fundamental Property assumes that $W$ is not omitted in $H$, so
there is an abstract execution $B$ with hardware realization $G$
witnessing that the dependency of the store $W$ on the loads
$\{R_i\}$ in $A$ is preserved in $H$.
We must show that some $R_i$ is not omitted and $R'_i$ is ordered
before $W'$ in $H$.
The proof splits into three cases.

First case: $W'$ lies in the initial period of $T$ in $H$.
During the initial period of the hardware executions, $H$ and $G$
behave identically and therefore $W'$ performs the same write in
both.
This contradicts the fact that $B$ and $G$ witness the preservation
of the dependency.

Second case: $W'$ lies in the common period of $T$ in $H$.
Since the action of $W'$ in $H$ is different from its action in $G$,
at least one of its inputs must differ between the two hardware
executions.
Therefore the source instruction for that input must behave
differently, and so must one of its sources, going back until we reach
a load instruction that obtains differing values in $H$ and $G$.
Then $W'$ depends on this load and so is ordered after it.
And since the load must lie in the common period of $H$, by the
definition above it must either be one of the
$R'_i$ or be ordered after one of them.
Therefore $W'$ is ordered after one of the $R'_i$ in $H$,
which certainly means that $R_i$ is not omitted.

@@

A corollary of this result is that if an implementation's compiler
either
(1)~uses per-thread analysis and treats atomic objects as quasi
volatile, or
(2)~uses global analysis and treats atomic objects as volatile,
then programs produced by that implementation will never exhibit
an OOTA cycle in which the semantic dependencies are preserved.
And since by the DP thesis, programs produced by
the implementation always preserve semantic dependencies,
these programs will never exhibit any OOTA cycles at all.
Thus the implementation will automatically comply with full C++, even
if it was designed only to comply with loose C++.

@@

Here we consider some general questions related to semantic dependency
and our definition of dependency preservation.
Additional information is available in P3064R2 (``How to Avoid OOTA Without
Really Trying'')~\cite{PaulEMcKenney2023P3046R2/OOTAwithoutTrying}.

\subsection{Defining Semantic Dependency}
\label{sec:Defining Semantic Dependency}

Some readers may object that our approach does not include any
rigorous definition of semantic dependency itself.
One could address this drawback by turning our DP thesis around
and using it instead as a \emph{definition} of semantic dependency.
In other words, a syntactic dependency could be considered to be semantic
if it is preserved by any valid loose C++
implementation, past, present, or future, real or imagined.
Of course this notion has its own problems, including that it is
extremely nonconstructive and impossible to apply in practice.
However it may be the best we can do with our current understanding
of computing systems.

Either way, regardless of how one wants to define semantic dependency,
we have shown that as long as the DP thesis continues to hold,
programs produced by an implementation of the right sort will never
exihibit OOTA.

@@

One can ask if we really need to require global-analysis compilers to
treat atomic objects as volatile; would our results still hold if they
merely treated them as quasi volatile?
Answer: The proof that preserved dependencies obey the Fundamental Property
would go through as before,
but the DP thesis would no longer be valid.
Here's an example showing how it would fail.

Consider the Simple OOTA program shown in
Listing~\ref{lst:Simple OOTA}.
A loose C++ compiler using global analysis and treating \co{x} and \co{y} as
quasi-volatile objects could omit the two loads, replacing them in the
machine code with simple assignments ``\co{r1 = 42}'' and
``\co{r2 = 42}'', as shown in
Section~\ref{sec:Global Optimization Can Destroy Dependencies}.
This would be a valid transformation,
but it would destroy the dependencies in T0 and T1 rather than preserving them,
even though those dependencies are considered to be semantic.

This fact is intuitively obvious, but let us describe in detail
exactly how the preservation definition would fail.
Recall that according to the definition, in order to be preserved in an execution
a syntactic dependency must have a witness,
another execution in which the store acts differently.
But this transformed program has no other hardware executions; every time it
runs it will store 42 to both \co{x} and \co{y}.
(Keep in mind also that since the atomic objects are not treated as
volatile, they are not subject to unspecified interference when the
program runs in a nonbenign environment.)

This unintuitive behavior could not occur if the two loads were not
omitted in the transformed program.
In fact, the definition of dependency preservation might remain
strong enough to justify the DP thesis
if the requirement for global-analysis compilers
were weakened, if the compiler were allowed to treat atomic objects as
quasi volatile except that it was not allowed to omit accesses to them.
This is a possible topic for future research.

@@

Part of our demonstration of the Fundamental Property for preserved
dependencies relies on the fact, stated in
Section~\ref{sec:Abstract Executions},
that an abstract execution of a thread is entirely determined by the
values obtained for its loads.
But when we compare abstract executions of the same thread in two
different programs, this may no longer be entirely true owing to the
effect of differing memory layouts.

@@

To rule out such pathological counterexamples we should require that
in a witness to a preserved dependency, the addresses of all the
objects and functions referred to in the thread $T$ are the same as
in the original execution.
This is a very technical restriction but there are occasions when
the issue might realistically arise, such as when computing a hash
value based on an object's address.

@@

To demonstrate that the dependency in the original code is preserved,
a suitable witness would have
to include a hardware realization of an abstract execution in which
\co{r1} and \co{r2} are zero and \co{r3} and \co{r4} are one.
But there are no such hardware realizations with the
machine code indicated above!
Since \co{r1} is different from both \co{r3} and \co{r4}, the \co{r2}
load will be merged with the \co{r3} load and so \co{r2} will
necessarily be one, not zero.
Thus the dependency is not preserved, which leads one to question
whether it really was semantic.

Although this result is unexpected, and while we don't recommend this sort of
code transformation, we cannot say that this conclusion is definitely
wrong, because one's intuitive notions of semantic dependency are not
always clear in cases involving multiple loads from the same variable.
% Can we instead say that transforming the code might require that the
% witness also be transformed?
In the unlikely event that this transformation proves useful, there
will be a community process that determines whether the value of this
transformation is worth the added conceptual load on developers.
Until that time, pure mathematics cannot help us simply because it is
impossible to say whether or not there is a semantic dependency from
\co{y} to \co{z} without reference to a specific implementation.

@@

Additional future work could:
(1)~Extend these results to some classes of interpreters and just-in-time
compilers (JITs) on
one hand, and to link-time optimization (LTO) on the other;
(2)~Relax quasi-volatile semantics to permit limited load speculation
(for example, hoisting loads out of the bodies of conditionals);
(3)~Consider various classes of global analysis, such as
demonstrating that a given expression will always evaluate to a constant;
(4)~Examine the possibility of ``flattening'' optimizations that combine
multiple threads into one;
(5)~Explore more detailed justifications for the Dependency Preservation thesis;
(6)~Include the effects of input and output or other operations which
can vary from one execution to another; and
(7)~Delineate more precisely the limits of permissible behavior for
quasi-volatile object accesses.
This last item may well shed additional light on the dangers of
optimizations involving non-volatile atomic operations beyond the cautions
outlined in
Section~\ref{sec:Inventing Atomic Loads Can Cause Errors and
Destroy Semantic Dependencies}.

@@

This paper focused solely on compilers, primarily those that do only
per-thread analysis and optimization.
It introduced the notion of quasi-volatile object accesses and gave
a definition for dependency preservation by a C++ implementation.
It formulated a Fundamental Property of semantic dependency along with
a Dependency Preservation thesis, and using the thesis proved
that implementations subject to some very mild restrictions on how
they treat atomic object accesses do have this property.
Finally, it demonstrated that satisfying the Fundamental Property
guarantees an implementation cannot give rise to OOTA executions.
